The impact of gas bulk rotation on the Lyman $\alpha$ line

### 2014-02-20 paper/ms.tex
mi version

\newcommand{\avg}[1]{\langle{#1}\rangle}  
\newcommand{\nscatt}{\langle N_{\rm  scatt}\rangle}
\newcommand{\ly}{{\ifmmode{{\rm Ly}\alpha~}\else{Ly$\alpha$~}\fi}}
\newcommand{\hMpc}{{\ifmmode{h^{-1}{\rm Mpc}}\else{$h^{-1}$Mpc }\fi}}   
\newcommand{\hGpc}{{\ifmmode{h^{-1}{\rm Gpc}}\else{$h^{-1}$Gpc }\fi}}   
\newcommand{\hmpc}{{\ifmmode{h^{-1}{\rm Mpc}}\else{$h^{-1}$Mpc }\fi}}  
\newcommand{\hkpc}{{\ifmmode{h^{-1}{\rm kpc}}\else{$h^{-1}$kpc }\fi}}  
\newcommand{\hMsun}{{\ifmmode{h^{-1}{\rm
        {M_{\odot}}}}\else{$h^{-1}{\rm{M_{\odot}}}$}\fi}}   
\newcommand{\hmsun}{{\ifmmode{h^{-1}{\rm
        {M_{\odot}}}}\else{$h^{-1}{\rm{M_{\odot}}}$}\fi}}   
\newcommand{\Msun}{{\ifmmode{{\rm {M_{\odot}}}}\else{${\rm{M_{\odot}}}$}\fi}}  
\newcommand{\msun}{{\ifmmode{{\rm {M_{\odot}}}}\else{${\rm{M_{\odot}}}$}\fi}}  
\newcommand{\lya}{{Lyman$\alpha$~}}
\newcommand{\clara}{{\texttt{CLARA}}~}
\newcommand{\rand}{{\ifmmode{{\mathcal{R}}}\else{${\mathcal{R}}$ }\fi}}  
\newcommand{\hs}{{\hspace{1mm}}}  
\newcommand{\kms}{{\ifmmode{{\mathrm{\,km\ s}^{-1}}}\else{\,km~s$^{-1}$}\fi}}

@@

We present results of radiative transfer calculations to measure the
impact of gas bulk rotation on the morphology of the Lyman $\alpha$
emission line in distant galaxies. We model a galaxy as a sphere with
a homogeneous mixture of dust and hydrogen at a constant
temperature. These spheres have a solid-body rotation with maximum
velocities in the range $0-300$ \kms and neutral hydrogen optical
depths in the range $\tau_{\rm H}=10^{5}-10^{7}$. We consider cases 
of a single central \lya source, and a uniform distribution of sources. 
Our main finding are that the escape fraction of \lya photons, and
the intensity at the line center increase with rotational velocity. 
For homogeneously distributed sources the line becomes single peaked 
at rotational velocities larger than the line width in the static case.
For radiation sources located off-center, the line morphology presents 
a range of single, double and triple peaks. We show how these results 
are useful to interpret recent spectroscopic results of distant 
$z\sim 2-3$ star forming galaxies.

@@

Performing that study is the main goal of this paper. We investigate for the
first time the impact of rotation on the morphology of the \ly
line. We focus on a simplified system: a spherical gas cloud with
homogeneous density and solid body rotation, to study the line
morphology and the escape fraction in the presence of dust. We base
our work on two independent Monte Carlo based radiative transfer codes
CLARA \citep{CLARA} and \citep{DijkstraKramer}.

This paper is structured as follows: In \S \ref{sec:implementation} we
present the implementation of bulk rotation into the Monte Carlo
codes, paying special attention to coordinate definitions. We also
present a short review of how the \ly radiative transfer codes work
and list the different physical parameters in the simulated grid of
models. In the next \S \ref{sec:results} we present the results of the
simulations, with special detail to quantities in the line that show a
clear evolution as a function of the sphere rotational velocity. In \S
\ref{sec:discussion} we discuss the implications of our results in the
interpretation of LAEs observations. In the last section we present
our conclusions.

@@

Describing the kinematics of gas rotation in all generality is a
complex task, specially at high redshift where there is still missing
a thorough observational account of rotation in galaxies beyond
$z>1.0$. Event at low red-shifts there is a great
variation in the shape of the rotation curve as observed in HI
emission as a function of the distance to the galaxy center. However
there are two recurrent features. First, in the
central galactic region the velocity increases proportional to the radius,
following a solid rotation behavior. Second, beyond a certain radius
the rotation curve tends to flatten.

An ab-initio description of realistic rotation curves in simulations
depends on having access to the dynamic evolution of all mass components
in the galaxy: stars, gas and dark matter. Such level of realism is
extremely complex to achieve, specially if one wants to get a
systematic description based on statistics of simulated objects.

@@

For each photon in the simulation we have its initial position inside
the sphere, direction of propagation $\hat{k}_{in}$ and reduced
frequency $x_{in}$. The photon's propagation stops once they cross the
surface of the sphere. At this point we store the position, the outgoing direction
of propagation $\hat{k}_{out}$ and the reduced frequency $x_{out}$. We
define the angle $\Theta$ by $\cos\Theta = \hat{k}\cdot
\hat{k}_{out}$, that is the polar angle of the outgoing photon with
respect to the $z$ axis. Following \cite{Zheng2013} we make the study
of the anisotropic emission in terms of this angle.

@@

The codes follow the individual scatterings of \ly photons as it
travels through a 3D distribution of neutral Hydrogen. At each
scattering the frequency of the photon (in the laboratory frame) and
its direction of propagation change. This change in frequency is due
to the peculiar velocities of the Hydrogen atom that absorbed and
re-emitted the photon. If dust is present, the photon can interact
either with a Hydrogen atom or dust grain. In the case of a dust
interaction the photon can be either absorbed or scattered, this
probability is encoded in the dust albedo, $A$, which we chose to be
$1/2$. In order to obtain accurate values for the escape fraction of
photons in the presence of dust, we do not use any accelerating
mechanism in the radiative transfer. Once the photons escape the gas
distribution we store their direction at their direction of
propagation and frequency at their last scattering.

@@

The gas is completely defined by its geometry (i.e. sphere or slab),
temperature $T$, Hydrogen optical depth $\tau_{\rm H}$, dust optical
depth $\tau_{\rm a}$ and the bulk velocity field $\vec{v}$. Here we
treat the gas as homogeneous in density ($\tau_{\rm H}$, $\tau_{\rm
a}$) and temperature.

\begin{figure*}
\begin{center}
  \includegraphics[width=0.8\textwidth]{f1.eps}
\end{center}
\caption{Shape of the \ly line for
    different velocities rotational velocities for spherical
    distributions in all the models. The continuous (dashed) line represents the 
    central (homogeneous) source distributions.
    \label{fig:differentvelocities}}  
\end{figure*}

@@

In the Monte Carlo calculations we follow the propagation of $N_{\gamma}=10^5$
numerical photons through different spherical galaxies. For each galaxy
we vary at least one of the following parameters: the maximum
rotational velocity $V_{\rm max}$, the hydrogen optical depth $\tau_{H}$,
the dust optical depth $\tau_{a}$ and the initial distribution of photons
with respect to the gas. There are $60$ models initial combining all
variations of the input parameters. Table \ref{table:models}
summarizes the models.

@@

The central result of this paper is summarized in Figure
\ref{fig:differentvelocities}. It shows the evident
impact of rotation on the morphology of the emergent \ly line.
The panel focus on the results for all of our symmetric models
 $\tau_{H}=(10^{5}, 10^{6}, 10^{7}$), showing that the
influence of rotation is present both when the photons are either
homogeneously or centrally initialized over the gas volume.

In the next subsections we characterize the line morphology by the
full width at half maximum (FWHM) and the location of the peak
maxima. In order to interpret the morphological changes in the line we
also report the median number of scatter for each \ly photon in the
simulation. For the models where dust is included we measure the 
escape fraction as a function of rotational velocity. Finally, we make
an estimate of the anisotropic emission of the models in comparison
with static spheres.

@@

There are three clear effects on the line's morphology as the
rotational velocity increases. First, the line
broadens; second, the double peaks reduce their intensity; and third,
the intensity at the line center rises. The two last effects give the
impression that the double peaks are merger into one at high rotational
velocities, a result that is evident for the homogeneously distributed
sources as shown in the dashed lines of Figure \ref{fig:differentvelocities}.

We use the full width at half maximum (FWHM) to quantify the line
broadening. Figure \ref{fig:widthvsvelocity} shows how the FWHM increases
with rotational velocity both for the central an homogeneous source
distributions.  We parametrize the dependency of the line width as
\begin{equation}
 {\rm FWHM}^2 = {\rm FWHM}_{0}^2 + V_{\rm max}^2/\lambda^2,
\label{eq:fwhm}
\end{equation}
%
where FWHM$_{0}$ is the
velocity width in the static case and $\lambda$ is a positive scalar
to be determined as a fit to the data. With this test we want to know
to what extent the new velocity width can be expressed as a quadratic
sum of the two relevant velocities in the problem. We fit
simultaneously all the points corresponding to central and homogeneous
models to find $\lambda_{\rm c} = 3.9 \pm 0.041$ and $\lambda_{\rm h}=
1.9\pm 0.062$ respectively.

In order to quantify the transition to a single peak we measure the
position for the peak maxima as a function of the rotational
velocity. Our results are presented in Figure
\ref{fig:maximumsvsvelocity} where it is clear that for central
sources there are always two peaks with fixed positions as the rotational
velocity changes. However, in the case of homogeneously emitted
sources the maxima position remain close to constant until some
velocity threshold the line becomes single peaked with $x_{\rm m}=0$
\kms. By inspection we find that in all cases the transition has
occurred when the rotational velocity is larger than half width.

One possible explanation for the emergence of the single peak in the
homogeneous source systems is that some photons close to the surface
(a sort of photosphere) can escape with a low number of scatterings,
allowing them to stay close to the line's center. Increasing the
rotational velocity $V_{\rm max}$ reduces the optical depth making the
photosphere region effectively larger, increasing the number of
photons escaping close to the line's center.

However, for the central emission the transition to a single peak is
never completed in the range of explored parameters. The absence
of a single peak phase could be partially explained by the absence of a
photosphere, in this case the average number of scatterings to escape
should remain close to constant as the rotational velocity
increases. The rise in intensity at the line's center could instead
mean that the scattering in a medium with bulk motion are inefficient
in driving photons outside that frequency region.

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.45\textwidth]{f5_1.eps}
  \includegraphics[width=0.45\textwidth]{f5_2.eps}
  \includegraphics[width=0.45\textwidth]{f5_3.eps}
  \includegraphics[width=0.45\textwidth]{f5_4.eps}    
\end{center}
    \caption{2D histogram of $N_{\rm scatt}$ vs $x$. The upper (lower) panels
      show the homogeneous (central) source distribution. Left panels 
      corresponds to the static case and right panels corresponds to
      $V_{\rm max}=300\kms$. The color scale is logarithmic on the
      number of photons with given values of $N_{\rm scatt}$ and
      $x$. \label{fig:Nscatt2D}}   
\end{figure*}

@@

In Figure \ref{fig:Nscatt} we show the average number of scatterings
$\langle N_{\rm scatt}\rangle$ as a function of the rotational velocity
$V_{\rm max}$. For the central source distribution the average number of
scatterings $\langle N_{\rm   scatt}\rangle$ changes less than $0.5\%$
for different velocities. In this set-up the number of scatterings is
proportional to the optical depth, with $\langle N_{\rm
  scatt}\rangle= (1.50, 1.00, 0.92)\tau_{\rm   H}$ for optical depth
values of $\tau_{\rm H} = (10^{5}, 10^{6}, 10^{7})$ respectively and
$V_{\mathrm{max}}=300$ \kms.

From Figure \ref{fig:Nscatt} for the homogeneous
distribution there is a decrease of $\langle N_{\rm
  scatt}\rangle$ as the $V_{\rm max}$ increases. For instance, for
$\tau_{\rm H}=10^5$ the average number of scatterings decreases by
$61\%$ at $V_{\rm max}=300\kms$ in comparison to the static case.

For the homogeneous static sphere we find that $\langle N_{\rm
  scatt}\rangle= (0.99, 0.59, 0.51)\tau_{\rm   H}$, this
represents factors of $(0.66, 0.59, 0.55)$ lower than the centrally
emitted photons. While for the homogeneous rotating sphere at 
$V_{\rm{max}}=300\kms$ we find that $\langle N_{\rm
  scatt}\rangle=(0.38, 0.34, 0.42)\tau_{\rm   H}$ which represents 
  factors of $(0.26, 0.34, 0.46)$ lower than the corresponding central
  case.  In contrast the analytic solution for the homogeneous
slab gives	 $0.72$ times less scatterings in the homogeneous case than
in the central one

In order to gain a deeper understanding of these results and explain
the results for the maxima in the line morphology we make 2D
histograms for the number of scatterings as a function of the outgoing
dimensionless frequency $x$. In Figure \ref{fig:Nscatt2D} we show
such histograms in the case $\tau_{\rm H}=10^5$ for the
static case and $V_{\rm max}=300$\kms. The upper (lower) panels show the
results for the homogeneous (central) source distribution. The color
scale is logarithmic in the number of photons at a certain value
$xN_{\rm scatt}$.

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.40\textwidth]{f6.eps}
  \includegraphics[width=0.40\textwidth]{f7.eps}
\end{center}
  \caption{Effects of rotation on the escape fraction. Left
    panel. Escape fraction as a function of rotational velocity. All
    these models have $\tau_{a}=1$. The continuous (dashed) lines
    correspond to central (homogeneous) models. Right panel. Escape
    fraction as a    function of the product $(a\tau_{\rm
      H})^{1/3}\tau_{A}$. The    analytic solution for the infinite
    slab is shown as a continuous    line. Different lines corresponds
    to different rotational velocities for the homogeneous models.     \label{fig:efvsv}} 
\end{figure*}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.90\textwidth]{f11.eps}
\end{center}
\caption{Triple peaked profiles for three different off-center positions
  $C=0.25R$ (doted line), $C=0.5R$ (dashed line) \& $C=0.75R$ (solid
  line) and for different rotational velocities. And $\theta$ represents the usual
  polar angle.
   \label{fig:3p_profiles}} 
\end{figure*}

@@

We also explore a different source of anisotropy by studying the effect
of radiation sources located in off-centered spheres with respect to
the gas distribution. Figure \ref{fig:OCspheres} shows the two
spherical regions we define to test for this effect. Each has a radius
of $0.5R$, where $R$ is the radius of the gaseous sphere. With an
observer located along the $\hat{j}$ direction, we pick two possible
positions to center the spheres along the $\hat{k}$ axis: $\pm C\hat{k}$. In
this configuration the emitting sphere centered on $+C\hat{k}$ moves
away from the observer and the sphere centered on $-C\hat{j}$ moves
towards the observer.

In this case the $\mu$ distribution is not homogeneous as shown in
Figure \ref{fig:muhisto}. The off-center emission shows a variation of flux at
the levels of $10\% - 28\%$ in intensity with the angle $\Theta$. In general 
models with a larger $\tau_{H}$ present a higher variation in the flux.
This variation implies that that the line is more intense in
the plane perpendicular to the rotation axis.

We also explore the effects of the off-centered emission in the line
morphology. Figure \ref{fig:OC_profiles} show the spectra for the two
off-centered emitting spheres located at $C=\pm0.5R$.  In this case we
only use photons that have outgoing directions aligned with the
observer i.e. $0.9 <\vert \mu \vert < 1.0$. We do not find any strong
asymmetry in the line morphology.

However, we find some cases where there is a possible signature of
three peaks. This is only present in some models ($\tau_{\rm H} = 10^6,
V=100$\kms and $\tau_{\rm H} = 10^7, V=300$\kms) and could be the result of
two competing effects: the rise in the line's center as the photons in
the photosphere escape in a single excursion and the broadening in the
two peaks as the photons closer from the center have to scatter to
escape. We examine in greater detail these cases in the next
sub-section.

@@

The asymmetries in Groups I, II  and III cannot be explained by
rotation. However, some galaxies in those groups have a high flux between
the two peaks; this could indicate the presence of rotation 
and \lya photons leaking though low column density holes, as in \citep{Behrens2014}.
 The same argument applies to galaxies in the group IV in which the flux
in between of the two peaks is high ($ \geq 70 \%$ of the intensity in
the highest peak). Finally, the triple peaked lines in Galaxies in
Group V can be easily reproduced by rotation with off-centered
emission given the right amount of asymmetry between the emitting
region and neutral hydrogen distribution.

@@

Our main result is that rotation clearly impacts the \ly line
morphology. Double peaked lines can make transitions into single
peaked lines as the rotational velocity increases. In the case of
off-centered emission result into triple peaked lines for certain
values of optical depth and rotational velocities.

@@

\item The line width increases with rotational velocity. The line with
  approximately following the functional for  ${\rm FWHM}^2 = {\rm FWHM}_{
    0}^2 + (V_{\rm max}/\lambda)^2$ where FWHM$_{0}$ indicates the line
  width for the static case. $\lambda$ is a constant determined from
  the radiative transfer results it is $\lambda_{\rm c}=3.9\pm0.041$ and
  $\lambda_{\rm h}=1.9\pm0.062$ for the central and homogeneous source
  distribution, respectively.

\item A single peaked line emerges for high rotational velocities in
  the case of homogeneously distributed sources. These cases occur when
  the rotational velocity is larger than FWHM$_0$.

@@

\item For homogeneous sources we find that the changes in the
line morphology are linked to the reduction in the average number of
scatterings. In the second part of \S \ref{sec:widthpeak} we show
that, as rotational velocity increases, a large fraction of photons
escape with less than $10$ scatterings. The average number of
scatterings decreases a $40\%$ compared to the static case. 
The reduction in the number of scatterings decreases the probability
for a photon to move away from the line's center. This is also
translated into an increase of the escape fraction by a factor of
$\sim 2$ with respect to the static situation.

\item The spatial anisotropy in the emission is always below the
  $10\%$ level deviation from static models.

\item For off-centered emission triple peaked profiles can emerge
  under two conditions: the emission occurs close to the sphere's
  surface and the rotational velocity is half the FWHM in the static
  case.   
\end{itemize}

>>>>

\newcommand{\avg}[1]{\langle{#1}\rangle}  
\newcommand{\nscatt}{\langle N_{\rm  scatt}\rangle}
\newcommand{\ly}{{\ifmmode{{\rm Ly}\alpha~}\else{Ly$\alpha$~}\fi}}
\newcommand{\hMpc}{{\ifmmode{h^{-1}{\rm Mpc}}\else{$h^{-1}$Mpc }\fi}}   
\newcommand{\hGpc}{{\ifmmode{h^{-1}{\rm Gpc}}\else{$h^{-1}$Gpc }\fi}}   
\newcommand{\hmpc}{{\ifmmode{h^{-1}{\rm Mpc}}\else{$h^{-1}$Mpc }\fi}}  
\newcommand{\hkpc}{{\ifmmode{h^{-1}{\rm kpc}}\else{$h^{-1}$kpc }\fi}}  
\newcommand{\hMsun}{{\ifmmode{h^{-1}{\rm
        {M_{\odot}}}}\else{$h^{-1}{\rm{M_{\odot}}}$}\fi}}   
\newcommand{\hmsun}{{\ifmmode{h^{-1}{\rm
        {M_{\odot}}}}\else{$h^{-1}{\rm{M_{\odot}}}$}\fi}}   
\newcommand{\Msun}{{\ifmmode{{\rm {M_{\odot}}}}\else{${\rm{M_{\odot}}}$}\fi}}  
\newcommand{\msun}{{\ifmmode{{\rm {M_{\odot}}}}\else{${\rm{M_{\odot}}}$}\fi}}  
\newcommand{\lya}{{Lyman $\alpha$~}}
\newcommand{\clara}{{\texttt{CLARA}}~}
\newcommand{\rand}{{\ifmmode{{\mathcal{R}}}\else{${\mathcal{R}}$ }\fi}}  
\newcommand{\hs}{{\hspace{1mm}}}  
\newcommand{\kms}{{\ifmmode{{\mathrm{\,km\ s}^{-1}}}\else{\,km~s$^{-1}$}\fi}}

@@

We present results of radiative transfer calculations to measure the
impact of gas bulk rotation on the morphology of the Lyman $\alpha$
emission line in distant galaxies. We model a galaxy as a sphere with
a homogeneous mixture of dust and hydrogen at a constant
temperature. These spheres have a solid-body rotation with maximum
velocities in the range $0-300$ \kms and neutral hydrogen optical
depths in the range $\tau_{\rm H}=10^{5}-10^{7}$. We consider cases 
of a single central \lya source, and a uniform distribution of sources. 
Our main finding are that the escape fraction of \lya photons, and
the intensity at the line center increases with rotational velocity. 
For homogeneously distributed sources the line becomes single peaked 
at rotational velocities larger than half the line width in the static case.
For radiation sources located off-center, the line morphology presents 
a range of single, double and triple peaks. We show how these results 
are useful to interpret recent spectroscopic results of distant 
$z\sim 2-3$ star forming galaxies.

@@

Performing that study is the main goal of this paper. We investigate for the
first time the impact of rotation on the morphology of the \ly
line. We focus on a simplified system: a spherical gas cloud with
homogeneous density and solid body rotation, to study the line
morphology and the escape fraction in the presence of dust. We base
our work on two independent Monte Carlo based radiative transfer codes
presented in \cite{CLARA} and \cite{DijkstraKramer}.

This paper is structured as follows: In \S \ref{sec:implementation} we
present the implementation of bulk rotation into the Monte Carlo
codes, paying special attention to coordinate definitions. We also
present a short review of how the \ly radiative transfer codes work
and list the different physical parameters in the simulated grid of
models. In \S \ref{sec:results} we present the results of the
simulations, with special detail to quantities that show a
clear evolution as a function of the sphere rotational velocity. In \S
\ref{sec:discussion} we discuss the implications of our results in the
interpretation of LAEs observations with high resolution
spectroscoy. In the last section we present our conclusions.

@@

Describing the kinematics of gas rotation in all generality is a
complex task, specially at high redshift where there is still missing
a thorough observational account of rotation in galaxies beyond
$z>1.0$. Even at low redshifts there is a great
variation in the shape of the rotation curve as observed in HI
emission as a function of the distance to the galaxy center. However
there are two recurrent features. First, in the
central galactic region the velocity increases proportional to the radius,
following a solid rotation behavior. Second, beyond a certain radius
the rotation curve tends to flatten.  An ab-initio description of
such realistic rotation curves in simulations depends on having access to
the dynamic evolution of all mass components in the galaxy: stars, gas
and dark matter. Such level of realism is extremely complex to
achieve, specially if one wants to get a systematic description based
on statistics of simulated objects.

@@

For each photon in the simulation we have its initial position inside
the sphere, direction of propagation $\hat{k}_{in}$ and reduced
frequency $x_{in}$. The photon's propagation stops once they cross the
surface of the sphere. At this point we store the position, the outgoing direction
of propagation $\hat{k}_{out}$ and the reduced frequency $x_{out}$. We
define the angle $\Theta$ by $\cos\Theta = \hat{k}_{\rm obs}\cdot
\hat{k}_{\rm out}$, that is the angle of the outgoing photon with
respect to the directions towards the observer. Following
\cite{Zheng2013} we make the study of the anisotropic emission in
terms of this angle.

@@

The codes follow the individual scatterings of \ly photons as they
travel through a 3D distribution of neutral Hydrogen. At each
scattering the frequency of the photon (in the laboratory frame) and
its direction of propagation change. This change in frequency is due
to the peculiar velocities of the Hydrogen atom that absorbed and
re-emitted the photon. If dust is present, the photon can interact
either with a Hydrogen atom or dust grain. In the case of a dust
interaction the photon can be either absorbed or scattered, this
probability is encoded in the dust albedo, $A$, which we chose to be
$1/2$. In order to obtain accurate values for the escape fraction of
photons in the presence of dust, we do not use any accelerating
mechanism in the radiative transfer. Once the photons escape the gas
distribution we store their direction at their direction of
propagation and frequency at their last scattering.

@@

The gas is completely defined by its geometry (i.e. sphere or slab),
temperature $T$, Hydrogen optical depth $\tau_{\rm H}$, dust optical
depth $\tau_{\rm a}$ and the bulk velocity field $\vec{v}$. Our codes
treat the gas as homogeneous in density ($\tau_{\rm H}$, $\tau_{\rm
  a}$) and temperature.

\begin{figure*}
\begin{center}
  \includegraphics[width=0.8\textwidth]{f1.eps}
\end{center}
\caption{Shape of the \ly line for
    different maximum rotational velocities. The continuous (dashed)
    line represents the central (homogeneous) source distributions. 
    \label{fig:differentvelocities}}  
\end{figure*}

@@

In the Monte Carlo calculations we follow the propagation of $N_{\gamma}=10^5$
numerical photons through different spherical galaxies. For each galaxy
we vary at least one of the following parameters: the maximum
rotational velocity $V_{\rm max}$, the hydrogen optical depth $\tau_{H}$,
the dust optical depth $\tau_{a}$ and the initial distribution of photons
with respect to the gas. There are $60$ models initial combining all
variations of the input parameters. Table \ref{table:models}
summarizes the different parameters.

@@

The central result of this paper is summarized in Figure
\ref{fig:differentvelocities}. It shows the clear
impact of rotation on the morphology of the emergent \ly line.
The panel focus on the results for all of our symmetric models
 $\tau_{H}=(10^{5}, 10^{6}, 10^{7}$), showing that the
influence of rotation is present both when the photons are either
homogeneously or centrally initialized over the gas volume.

In the next subsections we characterize the line morphology by the
full width at half maximum (FWHM) and the location of the peak
maxima. In order to interpret the morphological changes in the line we
also report the median number of scatter for each \ly photon in the
simulation. For the models where dust is included we measure the 
escape fraction as a function of rotational velocity. Finally, we make
an estimate of the impact of anisotropic emission.

@@

There are three clear effects on the line's morphology as the
rotational velocity increases. First, the line
broadens; second, the double peaks reduce their intensity; and third,
the intensity at the line center rises. The last two effects merge to give the
impression that the double peaks are merger into one at high rotational
velocities, a result that is evident for the homogeneously distributed
sources as shown in the dashed lines of Figure \ref{fig:differentvelocities}.

We use the full width at half maximum (FWHM) to quantify the line
broadening. Figure \ref{fig:widthvsvelocity} shows how the FWHM increases
with rotational velocity both for the central an homogeneous source
distributions.  We parametrize the dependency of the line width with
$V_{\rm max}$ as
\begin{equation}
 {\rm FWHM}^2 = {\rm FWHM}_{0}^2 + V_{\rm max}^2/\lambda^2,
\label{eq:fwhm}
\end{equation}
%
where FWHM$_{0}$ is the
velocity width in the static case and $\lambda$ is a positive scalar
to be determined as a fit to the data. With this test we want to know
to what extent the new velocity width can be expressed as a quadratic
sum of the two relevant velocities in the problem. We fit
simultaneously all the points corresponding to central and homogeneous
models to find $\lambda_{\rm c} = 3.89 \pm 0.04$ and $\lambda_{\rm h}=
1.98\pm 0.06$ respectively.

In order to quantify the transition to a single peak we measure the
position for the peak maxima as a function of the rotational
velocity. Our results are presented in Figure
\ref{fig:maximumsvsvelocity} where it is clear that for central
sources there are two peaks with fixed positions as the rotational
velocity changes, with the only exception of the model with $\tau_{\rm
  H}=10^6$ and $V_{\rm max}=300$\kms. However, in the case of
homogeneously emitted sources the maxima position remain fixed until
some velocity threshold is reachd and the line becomes single peaked
with $x_{\rm m}=0$. By inspection we find that in all cases the
transition has occurred when the rotational velocity is larger than
half the FWHM.

One possible explanation for the emergence of the single peak in the
homogeneous systems is that some photons close to the surface
(a sort of photosphere) can escape with a low number of scatterings,
allowing them to stay close to the line's center. Increasing the
rotational velocity $V_{\rm max}$ reduces the optical depth making the
photosphere region effectively larger, increasing the number of
photons escaping close to the line's center.

However, for the central emission the transition to a single peak is
almost never completed in the range of explored parameters. The absence
of a single peak phase could be partially explained by the absence of a
photosphere. In this case the average number of scatterings to escape
should remain close to constant as the rotational velocity
increases. The rise in intensity at the line's center could instead
mean that the scattering in a medium with bulk motion are inefficient
in driving photons outside that frequency region.

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.45\textwidth]{f5_1.eps}
  \includegraphics[width=0.45\textwidth]{f5_2.eps}
  \includegraphics[width=0.45\textwidth]{f5_3.eps}
  \includegraphics[width=0.45\textwidth]{f5_4.eps}    
\end{center}
    \caption{2D histogram of $N_{\rm scatt}$ vs $x$. The upper (lower) panels
      show the homogeneous (central) source distribution. Left panels 
      corresponds to the static case and right panels correspond to
      $V_{\rm max}=300\kms$. The color scale is logarithmic on the
      number of photons with given values of $N_{\rm scatt}$ and
      $x$. \label{fig:Nscatt2D}}   
\end{figure*}

@@

In Figure \ref{fig:Nscatt} we show the average number of scatterings
$\langle N_{\rm scatt}\rangle$ as a function of the rotational velocity
$V_{\rm max}$. For the central source distribution the average number of
scatterings $\langle N_{\rm   scatt}\rangle$ changes less than $0.5\%$
for different velocities. In this set-up the number of scatterings is
proportional to the optical depth, with $\langle N_{\rm
  scatt}\rangle= (1.50, 1.00, 0.92)\tau_{\rm   H}$ for optical depth
values of $\tau_{\rm H} = (10^{5}, 10^{6}, 10^{7})$ respectively.

In Figure \ref{fig:Nscatt} we can see the decrease of $\langle N_{\rm
  scatt}\rangle$ as $V_{\rm max}$ increases in the homogeneous
source distribution. For instance, for
$\tau_{\rm H}=10^5$ the average number of scatterings decreases by
$61\%$ at $V_{\rm max}=300\kms$ in comparison to the static case.

For the homogeneous static sphere we find that $\langle N_{\rm
  scatt}\rangle= (0.99, 0.59, 0.51)\tau_{\rm   H}$, this
represents factors of $(0.66, 0.59, 0.55)$ lower than the centrally
emitted photons. While for the homogeneous rotating sphere at 
$V_{\rm{max}}=300\kms$ we find that $\langle N_{\rm
  scatt}\rangle=(0.38, 0.34, 0.42)\tau_{\rm   H}$ which represents 
  factors of $(0.26, 0.34, 0.46)$ lower than the corresponding central
  case.

In order to gain a deeper understanding of these results and explain
the results for the maxima in the line morphology we make 2D
histograms for the number of scatterings as a function of the outgoing
dimensionless frequency $x$. In Figure \ref{fig:Nscatt2D} we show
such histograms in the case $\tau_{\rm H}=10^5$ for the
static case and $V_{\rm max}=300$\kms. The upper (lower) panels show the
results for the homogeneous (central) source distribution. The color
scale is logarithmic in the number of photons at a certain value
$x-N_{\rm scatt}$.

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.40\textwidth]{f6.eps}
  \includegraphics[width=0.40\textwidth]{f7.eps}
\end{center}
  \caption{Effects of rotation on the escape fraction. Left
    panel, escape fraction as a function of rotational velocity. All
    these models have $\tau_{a}=1$. The continuous (dashed) lines
    correspond to central (homogeneous) models. Right panel, escape
    fraction as a    function of the product $(a\tau_{\rm
      H})^{1/3}\tau_{A}$. The    analytic solution for the infinite
    slab is shown as a continuous    line. Different lines correspond
    to different rotational velocities for the homogeneous models.     
    \label{fig:efvsv}} 
\end{figure*}

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.90\textwidth]{f11.eps}
\end{center}
\caption{Triple peaked profiles for three different off-center positions
  $C=0.25R$ (doted line), $C=0.5R$ (dashed line) \& $C=0.75R$ (solid
  line) and for different rotational velocities. $\theta$ represents
  the position of the observer as mesured with the polar angle.
   \label{fig:3p_profiles}} 
\end{figure*}

@@

We also explore a different source of anisotropy by studying the effect
of radiation sources located in off-centered spheres with respect to
the gas distribution. Figure \ref{fig:OCspheres} shows the two
spherical regions we define to test for this effect. Each has a radius
of $0.5R$, where $R$ is the radius of the gaseous sphere. With an
observer located along the $\hat{j}$ direction, we pick two possible
positions to center the spheres along the $\hat{i}$ axis: $\pm C\hat{i}$. In
this configuration the emitting sphere centered on $+C\hat{i}$ moves
towards the observer and the sphere centered on $-C\hat{j}$ moves
away from the observer.

In this case the $\mu$ distribution is not homogeneous as it is shown in
Figure \ref{fig:muhisto}. The off-center emission shows a variation of flux at
the levels of $10\% - 28\%$ in intensity with the angle $\Theta$. In general 
models with a larger $\tau_{H}$ present a higher variation in the flux.

We also explore the effects of the off-centered emission in the line
morphology. Figure \ref{fig:OC_profiles} show the spectra for the two
off-centered emitting spheres located at $C=\pm0.5R$.  In this case we
only use photons that have outgoing directions aligned with the
observer i.e. $0.9 <\vert \mu \vert < 1.0$. We do not find any strong
asymmetry in the line morphology within the range of explored velocities.

However, we find some cases where there is a possible signature of
triple peaked emission. This is only present in some models
($\tau_{\rm H} = 10^6, V=100$\kms and $\tau_{\rm H} = 10^7,
V=300$\kms) and could be the result of two competing effects: the rise
in the line's center as the photons in the photosphere escape in a
single excursion and the broadening in the two peaks as the photons
closer from the center have to scatter to escape. We examine in
greater detail these cases in the next sub-section.

@@

The asymmetries in Groups I, II  and III cannot be explained by
rotation. However, some galaxies in those groups have a high flux between
the two peaks; this could indicate the presence of rotation 
and \ly photons leaking though low column density holes, as in \cite{Behrens2014}. The same argument applies to galaxies in the group IV in which the flux
in between of the two peaks is high ($ \geq 70 \%$ of the intensity in
the highest peak). Finally, the triple peaked lines in Galaxies in
Group V can be reproduced by rotation with off-centered
emission given the right amount of asymmetry between the emitting
region and neutral hydrogen distribution.

@@

Our main result is that rotation clearly impacts the \ly line
morphology. Double peaked lines can make transitions into single
peaked lines as the rotational velocity increases. In the case of
off-centered emission we obtain triple peaked lines for certain
values of optical depth and rotational velocities.

@@

\item The line width increases with rotational velocity. The line width
  approximately follows the functional for  ${\rm FWHM}^2 = {\rm FWHM}_{
    0}^2 + (V_{\rm max}/\lambda)^2$ where FWHM$_{0}$ indicates the line
  width for the static case and $\lambda$ is a constant determined from
  the radiative transfer results it is $\lambda_{\rm c}=3.89\pm0.04$ and
  $\lambda_{\rm h}=1.98\pm0.06$ for the central and homogeneous source
  distribution, respectively.

\item A single peaked line emerges for high rotational velocities in
  the case of homogeneously distributed sources. These cases occur when
  the rotational velocity is larger than half FWHM$_0$.

@@

\item For homogeneous sources we find that the changes in the
line morphology are linked to the reduction in the average number of
scatterings. In the second part of \S \ref{sec:widthpeak} we show
that, as rotational velocity increases, a large fraction of photons
escape with less than $10$ scatterings. The average number of
scatterings decreases a $40\%$ compared to the static case. 
This reduction in the number of scatterings decreases the probability
for a photon to move away from the line's center. This is also
translated into an increase of the escape fraction by a factor of
$\sim 2$ with respect to the static situation.

\item The spatial anisotropy induced by rotation in the emission is
  usually below the $10\%$ level.

\item For off-centered emission triple peaked profiles can emerge
  under two joint conditions: the emission occurs close to the sphere's
  surface and the rotational velocity is close to half the FWHM in the
  static case.   
\end{itemize}
